A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-05 tex/codatadp.tex
more

@@

\begin{abstract}
Codatatypes naturally capture potentially infinite data structures and
processes. We present a decision procedure that combines reasoning about
datatypes and codatatypes. The dual of the cyclicity rule for datatypes is a
bisimilarity rule that identifies equal values, also in the presence of cycles.
The procedure is complete for universal problems and is compatible with the
Nelson--Oppen method. It has been implemented in the latest version of CVC4, a
state-of-the-art SMT solver. An evaluation based on problems generated from
Isabelle theories demonstrates the potential of the procedure for both proving
and model finding.
\end{abstract}

@@

In this paper, we introduce a unified decision procedure for ground problems
involving both datatypes and codatatypes. The procedure is described abstractly
as a calculus (Section~\ref{sec:the-calculus}) and is designed to be composable
via the Nelson--Oppen method \cite{nelson-oppen-1979}.
For the datatype case, it
roughly follows the lines of Barrett et al.\ \cite{barrett-et-al-2007}.
To our knowledge, ours is the first decision procedure for codatatypes.

@@

Datatypes and codatatypes share many of the same properties, so it makes sense
to consider them together. There are, however, a number of important
differences:
%
\begin{itemize}
\item
\emph{No well-foundedness check is necessary for codatatypes.}
For example, the type of infinite sequences, or streams, over an
element type $\tau$
%
\[\keyw{codatatype}~\;\ty{stream}_{\,\tau} \,=\, \const{SCons}(\tau,\, \ty{stream}_{\,\tau})\]
is allowed; the corresponding datatype would be rejected as
non-well-founded.

@@

The proofs or soundness and completeness are included in the technical report
associated with this paper \cite{our-report}.
The procedure is implemented in the SMT solver CVC4 as rewriting and a theory solver
(Section~\ref{sec:the-theory-solver}).
It consists of about 2000 lines of \cpp{} code, among which 1600 are
shared between datatypes and codatatypes. The code is integrated in the
development version of the solver and is expected to be part of the CVC4~1.5 release.
%
An evaluation on %hand-crafted examples and on
problems generated from Isabelle theories using the Sledgehammer tool
demonstrates the usefulness of the approach (Section~\ref{sec:experimental-results}).

>>>>

\excludeversion{report}
\includeversion{paper}

\def\thewordpaper{paper}

\begin{report}
\def\thewordpaper{report}
\end{report}

@@

\begin{abstract}
Codatatypes naturally capture potentially infinite data structures and
processes. We present a decision procedure that combines reasoning about
datatypes and codatatypes. The dual of the cyclicity rule for datatypes is a
bisimilarity rule that identifies equal values, also in the presence of cycles.
The procedure is complete for universal problems and is compatible with the
Nelson--Oppen method. It has been implemented in the latest version of CVC4, a
state-of-the-art SMT solver. An evaluation based on problems generated from
Isabelle theories demonstrates the potential of the procedure.
\end{abstract}

@@

In this \thewordpaper, we introduce a unified decision procedure for ground and
universal problems involving both datatypes and codatatypes. The procedure is
described abstractly as a calculus (Section~\ref{sec:the-calculus}) and is
designed to be composable via the Nelson--Oppen method \cite{nelson-oppen-1979}.
For the datatype case, it
roughly follows the lines of Barrett et al.\ \cite{barrett-et-al-2007}.
To our knowledge, ours is the first decision procedure for codatatypes.

@@

Datatypes and codatatypes share many of the same properties, so it makes sense
to consider them together. There are, however, at least three important
differences:
%
\begin{itemize}
\item
\emph{No well-foundedness check is necessary for codatatypes.}
For example, the type of infinite sequences, or streams, over an
element type $\tau$
%
\[\keyw{codatatype}~\;\ty{stream}_{\,\tau} \,=\, \const{SCons}(\tau,\, \ty{stream}_{\,\tau})\]
is allowed; the corresponding datatype would be rejected as
non-well-founded.

@@

\begin{paper}
The proofs of soundness and completeness are included in the technical report
associated with this paper \cite{our-report}.
\end{paper}%
The procedure is implemented in the SMT solver CVC4 as rewriting and a theory solver
(Section~\ref{sec:the-theory-solver}).
It consists of about 2000 lines of \cpp{} code, among which 1600 are
shared between datatypes and codatatypes. The code is integrated in the
development version of the solver and is expected to be part of the CVC4~1.5 release.
%
An evaluation on %hand-crafted examples and on
problems generated from Isabelle theories using the Sledgehammer tool
demonstrates the usefulness of the approach (Section~\ref{sec:experimental-results}).
